---
title: "Class 5: Data Visualization Lab"
author: "Yane Lee PID A17670350"
format: pdf
editor: visual
---

# Week 3 Data Visualization Lab

Install the package ggplot

```{r}
#install.packages("ggplot2")
library(ggplot2)
#View(cars)
```


#A quick base R plot - this is not ggplot

```{r}
plot(cars)
```


# Our first ggplot, we need data + aes + geoms

```{r}
ggplot(data=cars) +
  aes(x=speed, y=dist) +
  geom_point()

p <- ggplot(data=cars) + 
  aes(x=speed, y=dist) +
  geom_point()
```


# Add a line with geom with geom_line()

```{r}
p + geom_line()
```


# Add a trend line close to the data

```{r}
p + geom_smooth()

p + geom_smooth(method="lm") 
```

#---------------------------------------------------

# Read in our drug expression data

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```


# Q. How many genes are in this dataset?

```{r}
nrow(genes)
```


# Q. How many 'up' regulated genes?

```{r}
table( genes$State )
```


# Q. What fraction of total genes is up-regulated?

```{r}
round((table(genes$State) / nrow(genes)) * 100, 2)
```


# Let's make a first plot attempt

```{r}
g <- ggplot(data=genes) + aes(x=Condition1, y=Condition2, col=State) + geom_point()

g
```


# Add some color

```{r}
g + scale_color_manual(values=c("pink", "lightyellow", "lightblue")) +
  labs(title="Gene expression changes", x= "Control (no drug)", y= "Drug Treatment") +
  theme_bw()
```

