---
title: "Machine Learning 1"
author: "Yane Lee PID A17670350"
date: "2026-02-01"
output: pdf_document
---

# First up kmeans()

Demo of using kmeans() function in based R. First make up some data with a known structure.

```{r}
tmp <- c( rnorm(30, -3), rnorm(30, 3) )
tmp
x <- cbind(x=tmp, y=rev(tmp))
plot(x)
```

Now we have some made up data in `x` let's see how kmeans works with this data

```{r}
k <- kmeans(x, centers = 2, nstart = 20)
k
```

> Q. How many points are in each cluster?

```{r}
k$size
```

> Q. How do we get to the cluster membership/assignment.

```{r}
k$cluster
```

> Q. What about cluster centers?

```{r}
k$centers
```

Now we got to the main results. Let's use them to plot our data with the kmeans result

```{r}
plot (x, col=k$cluster)
points(k$centers, col="blue", pch=15)
```

## Now for hclust()

We will cluster the same data `x` with the `hclust()`. In this case `hclust()` requires a distance matrix as input.

```{r}
hc <- hclust( dist(x) ) 
hc
```

Let's plot our hclust result

```{r}
plot(hc)
```

To get our cluster membership vector we need to "cut" the tree with the `cutree()`

```{r}
grps <- cutree(hc, h=8)
grps
```

Now plot our data with the hclust() results.

```{r}
plot(x, col=grps)
```

# Principal Component Analysis (PCA)

## PCA of UK food data

Read data from website and try a few visualizations.

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url, row.names=1)
x
```

```{r}
cols <- rainbow(nrow(x))
barplot( as.matrix(x), col=cols )
```

```{r}
barplot( as.matrix(x), col=cols, beside = TRUE )
```

```{r}
pairs(x, col=cols)
```

PCA to the rescue!!
The main base R PCA function is called `prcomp()` and we will ened to give it the transpose of our input data!

```{r}
t(x)
pca <- prcomp( t(x) )
```

```{r}
summary(pca)
```


```{r}
attributes(pca)
```

To make our new PCA plot (a.k.a. PCA score plot) we access `pca$x`

```{r}
plot(pca$x[,1], pca$x[,2])
text(pca$x[,1], pca$x[,2], colnames(x))
```

Color up the plot

```{r}
country_cols <- c("orange", "red", "blue", "green")
plot(pca$x[,1], pca$x[,2], xlab="PC1", ylab="PC2")
text(pca$x[,1], pca$x[,2], colnames(x), col=country_cols)
```


## PCA of RNA-Seq data

Read in data from website

```{r}
url2 <- "https://tinyurl.com/expression-CSV"
rna.data <- read.csv(url2, row.names=1)
head(rna.data)
```

```{r}
pca <- prcomp( t(rna.data) )
summary(pca)
```

Do our PCA plot of this RNA-Seq data

```{r}
plot(pca$x[,1], pca$x[,2], xlab="PC1", ylab="PC2")
```

```{r}
gene_cols <- c("lightpink", "skyblue", "yellow", "purple", "red", "blue", "orange", "lightgreen", "lavender")
plot(pca$x[,1], pca$x[,2], xlab="PC1", ylab="PC2")
text(pca$x[,1], pca$x[,2], colnames(rna.data), col=gene_cols)
```
